**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-08-05 paper/manuscript.Rmd
update figure colors,layout,order

---
bibliography: references.bib
link-citations: true
output:
  pdf_document:
    keep_tex: true
  word_document: default
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

@@

Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quantified how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or carcinoma) using five machine learning methods with the Mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p \< 0.05) but not family level (p = 0.06). These results were consistent for ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels). The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 2). While the resolution provided by ASVs is useful in

@@

@@

These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool. In general, we found that finer resolution (e.g. ASVs) does not add additional sensitivity to model predicition and at the ASV level actually impedes model performance due to the sparsity of shared taxa.

@@

![](../exploratory/figures/rf_with_stats.png){width="75%"}

**Figure 1: Random Forest Model Performance with Significance.** Boxplots of AUROC values on the test dataset for 100 seeds predicting SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which is equivalent to random classification. Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another.

@@

![](../exploratory/figures/bin_prevalence.png){height="50%"}\
**Supplemental Figure 2: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.

>>>>

---
bibliography: references.bib
link-citations: true
csl: msystems.csl
output:
  pdf_document:
    keep_tex: true
  word_document: default
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{multirow}
---

@@

dada2_rf_pvals <- read_csv("../data/dada2/analysis/dada2_pvalues_by_model.csv") %>% 
  filter(model == "rf")

@@

```{r pvals}
rf_pvals <- read_csv("../analysis/pvalues_by_model.csv",
                     col_types = cols(.default = col_character(),
                                      p_value = col_double())) %>% 
  filter(model == "rf") %>% 
  select(-model) %>% 
  pivot_wider(names_from = "level2",values_from = "p_value") %>% 
  column_to_rownames("level1")

rf_notsig <- read_csv("../analysis/pvalues_by_model.csv",
                     col_types = cols(.default = col_character(),
                                      p_value = col_double())) %>% 
  filter(model == "rf") %>% 
  filter(p_value >= 0.05)
```

Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quantified how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or carcinoma) using five machine learning methods with the Mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU (p \< 0.01), but borderline equivalent to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels. These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.

While AUROC values between models is a useful way to compare the overall model performance across all thresholds, they can be misleading [@lobo2008]. Depending on the intended implementation of the model, one may want to optimize the true-positive rate(or sensitivity) over the false-postivite rate (or specificity). 
To further compare the model performance across taxonomic levels we compared the sensitivity of the models

-   on average, how many positives are missed with ASV?

One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g. phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls. At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. Examination of the prevalence of taxa in samples at each level supports this idea. A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels. The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 3). While the resolution provided by ASVs is useful in

@@

Mikropml{}

@@

These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool. In general, we found that finer resolution (e.g. ASVs) does not add additional sensitivity to model prediction and at the ASV level actually impedes model performance due to the sparsity of shared taxa.

@@

```{r fig1}
#![](../exploratory/figures/rf_with_stats.png){width="50%"} ![](../exploratory/figures/rf_with_stats.png){width="50%"}
# library(cowplot)
# 
# p1 <- ggdraw + draw_image("../exploratory/figures/rf_with_stats.png", width = 0.5) 
# p2 <- ggdraw + draw_image("../exploratory/figures/rf_with_stats.png", width = 0.5, x = 0.5)
# 
# plot_grid(p1,p1,labels = c('A', 'B'), label_size = 12)

library(png)
library(grid)
library(gridExtra)
img1 <-  rasterGrob(as.raster(readPNG("../exploratory/figures/rf_with_stats.png")), interpolate = FALSE)
img2 <-  rasterGrob(as.raster(readPNG("../exploratory/figures/sens_90spec.png")), interpolate = FALSE)
grid.arrange(arrangeGrob(img1, left = textGrob("A)", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))),
             arrangeGrob(img2, left = textGrob("B)", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))), 
             ncol = 2)
```

**Figure 1: Random Forest Model Performance.** **A)** Boxplots with points of area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which is equivalent to random classification. Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another. **B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations. Black points denote the median and the lines denote the IQR. The letters W through Z denote the significance groups.

@@

![](../exploratory/figures/roc_by_level.png){height="50%"}

**Supplemental Figure 2: Averaged ROC curves** ROC curves averaged across the 100 iterations of the model. The shaded region represents the standard deviation form the mean.

![](../exploratory/figures/bin_prevalence.png){height="45%"}\
**Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.

![](../exploratory/figures/top10_imp.png){height="85%"}\
**Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUC when the feature is permuted.
